---
title: Crops, Europe, and {gt}
description: Creating a standardised {gt} table template to show top Countries by % of crop yielded
author: Duncan
date: '2020-11-08'
slug: crops-europe-and-gt
categories: [R, tidytuesday, tables, scraping]
tags: [tidytuesday, visualisation, scraping]
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(hrbrthemes)
library(gt)
theme_set(theme_ipsum_rc())

# tuesdata <- tidytuesdayR::tt_load('2020-09-01')
tuesdata <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-01/key_crop_yields.csv')

key_crops <- 
  #tuesdata$key_crop_yields %>%
  tuesdata %>%
  janitor::clean_names() %>% 
  rename(country = entity) %>%
  mutate(decade = 10*(year %/% 10))
```

Recently (or prior to US election craze) there seems to have been a bit of focus on tables in the R community on Twitter. 
<!--more-->

The `gt` package in particular, which was released by the amazing RStudio team, seems to have been getting lots of attention. There have been several TidyTuesday entries where people have used tables with the most notable (to me, from memory) entries by [Thomas Mock](https://themockup.blog/) who has written a few tutorials. Based on those tutorials I wanted to have a look myself. 


Tables are a necessity where I work to represent big piles of web data that can't be visualised easily and one of the issues we run into is that everyone uses different formats. Even the same person, based on what week they're doing something, will produce something completely different. `gt` seems to address this with a theme function so you can have cool consistent tables like the BBC or FiveThirtyEight, so I'll be trying that.

For the data I wanted to do something simple: look at [global crop yields](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-01/readme.md) and focus on European countries to see which is the top exporter for each crop throughout the decades. The below has several chunks stripped out, full code can be found on [GitHub](https://github.com/duncip/blog-posts/blob/master/2020-11-08-crops-europe-and-gt.Rmd)!


```{r}
key_crops %>%
  distinct(country) %>% nrow()
```

### Scraping and readying data 

With 249 countries there's a lot to process. Instead, I just want to focus on Europe. Because I don't want to put together a list manually, I'll use the [Wiki page](https://en.wikipedia.org/wiki/List_of_European_countries_by_population) to get a quick list and also to extract the flag icons for the table!

```{r include = FALSE}
library(rvest)
```

```{r}
url <- "https://en.wikipedia.org/wiki/List_of_European_countries_by_population"
wiki_scrape <- 
  url %>%
  read_html() %>%
  html_nodes(xpath = '//*[@id="mw-content-text"]/div[1]/table') 

wiki_table <-
  wiki_scrape %>%
  html_table(fill = TRUE)

names(wiki_table[[1]]) <- c("rank", "country", "population", "estimate", "date", "region", "source")
wiki_table[[1]]$country <- gsub('\\[.*?\\]', '', wiki_table[[1]]$country)
eu_list <- as_tibble(wiki_table[[1]])[-1,]

head(eu_list, 4)
```

Now that we have the list, I'll extract the flags separately. I also need to fix the URL as `gt` requires you to put the full url rather than the relative one:

```{r}
flags <- 
  wiki_scrape %>%
  html_nodes("img") %>%
  html_attr("src") %>%
  as_tibble() %>%
  rename("flag_url" = value) %>%
  mutate(country = str_extract(flag_url, "(?<=of_).+(?=.svg/)"),
         country = case_when(
           str_detect(country, "Kingdom") ~ "United Kingdom",
           str_detect(country, "Netherlands") ~ "Netherlands",
           str_detect(country, "Belgium") ~ "Belgium",
           TRUE ~ country
         ),
         flag_url = str_replace(flag_url, "\\/\\/", "https://"))

head(flags, 4)
```

With the list of European countries and flag URLs sorted, I had a quick look through the data and found that no countries in Europe produce **cocoa beans** or **cassava** so I've filtered both these out. Now we can have a quick look at recurring countries.

```{r include=FALSE}
eu_crop_data <- 
  key_crops %>%
  filter(country %in% eu_list$country) %>%
  group_by(country, code, decade) %>%
  summarise_at(vars("wheat_tonnes_per_hectare":"bananas_tonnes_per_hectare"), mean, na.rm = TRUE) %>%
  pivot_longer(cols = "wheat_tonnes_per_hectare":"bananas_tonnes_per_hectare", names_to = "crop", values_to = "yield") %>%
  filter(!is.na(crop)) %>%
  mutate(yield = if_else(is.na(yield), 0, yield),
         crop = str_replace(crop, "_tonnes_per_hectare", "")) %>%
  ungroup() %>%
  filter(!crop %in% c("cocoa_beans", "cassava"))
```


```{r echo = FALSE}
eu_crop_data %>%
  group_by(decade, crop) %>%
  mutate(decade_crop_pct = ifelse(yield == 0, 0, yield / sum(yield))) %>%
  ungroup() %>%
  group_by(decade, crop) %>%
  slice_max(decade_crop_pct, n = 1) %>%
  ungroup() %>%
  ggplot(aes(decade, decade_crop_pct, fill = country)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format()) +
  viridis::scale_fill_viridis(discrete = TRUE) +
  hrbrthemes::theme_ipsum_rc() +
  labs(x = NULL, y = NULL,
       title = "Quick look at top European countries",
       subtitle = "Based on yield %") +
  theme(plot.title.position = "plot",
        plot.title = element_text(size = 12),
        plot.subtitle = element_text(size = 9),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        legend.title = element_blank()) +
  facet_wrap(~crop, scales = "free") 
```

## Table time!

In terms of the data, I want to use percentages rather than actual yield so the numbers are comparable. I'll also simply slice the top for each crop, every decade, so we have the most productive countries only. This is the data I'll use for the first table:

```{r}
top_eu_table_data <-
  eu_crop_data %>%
  group_by(decade, crop) %>%
  mutate(decade_crop_pct = ifelse(yield == 0, 0, yield / sum(yield))) %>%
  ungroup() %>%
  group_by(decade, crop) %>%
  slice_max(decade_crop_pct, n = 1) %>%
  ungroup() %>%
  left_join(flags, by = "country") %>%
  select(-country)
```

To show you what the table would look like with any alterations at all, here's the first few lines..

```{r echo = FALSE}
top_eu_table_data %>%
  head(3) %>%
  gt()
```

&nbsp;

The way I defined the function is by altering the table one line at a time. I then decided what I could reuse for future tables and put it in a function. For example, I started with the above table and slowly added things like `opt_all_caps()` and got it to look how I wanted. Once that was done, I cut and pasted the reusable code into a function. My final function looks like this:

```{r}
gt_theme_cropc <- function(data,...) {
  data %>%
  opt_all_caps()  %>%
  opt_table_font(
    font = list(
      google_font("Roboto Condensed"),
      default_fonts()
    )
  ) %>%
    tab_style(
      style = cell_borders(
        sides = "bottom", color = "transparent", weight = px(2)
      ),
      locations = cells_body(
        columns = TRUE,
        rows = nrow(data$`_data`)
      )
    )  %>% 
    text_transform(                             
    locations = cells_body(
      contains("flag")
      ),
    fn = function(x) {
      web_image(
        url = x,
        height = 10
      )
    }
  ) %>%
    tab_style(
    style = list(
      cell_text(weight = "bold")
    ), 
    locations = cells_body(
      columns = vars(crop)
    )
  ) %>%
    cols_label(
    crop = "Crop"
    ) %>%
  tab_source_note(
    source_note = md("**DATA**:   [TidyTuesday wk 36](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-01/readme.md) <br>**TABLE**:    Duncan P")
  ) %>%
  tab_options(
    column_labels.background.color = "white",
    table.border.top.width = px(3),
    table.border.top.color = "transparent",
    table.border.bottom.color = "transparent",
    table.border.bottom.width = px(3),
    column_labels.border.top.width = px(3),
    column_labels.border.top.color = "transparent",
    column_labels.border.bottom.width = px(3),
    column_labels.border.bottom.color = "black",
    data_row.padding = px(3),
    source_notes.font.size = 8,
    table.font.size = 12,
    heading.align = "left",
    table.align = "left",
    ...
  ) 
}
```

### Let's see what the table looks like

```{r}
top_eu_table_data %>%
  select(code, decade, decade_crop_pct, flag_url, crop) %>%
  rename("country" = code) %>% 
  pivot_wider(names_from = decade, values_from = c(country, flag_url, decade_crop_pct)) %>% 
  select(crop, flag_url_1960, contains("1960"), flag_url_1970, contains("1970"), 
         flag_url_1980, contains("1980"), flag_url_1990, contains("1990"), 
         flag_url_2000, contains("2000"), flag_url_2010, contains("2010")) %>%
  gt() %>%
  fmt_percent(
    columns = contains("pct"),
    decimals = 0
  ) %>%
  tab_header(
    title = md("**Top exporter by crop for European countries through the decades**"),
    subtitle = md("% represents proportion of **total crop** export that decade")
  ) %>%
    cols_label(
    flag_url_1960 = "",
    flag_url_1970 = "",
    flag_url_1980 = "",
    flag_url_1990 = "",
    flag_url_2000 = "",
    flag_url_2010 = "",
    decade_crop_pct_1960 = "",
    decade_crop_pct_1970 = "",
    decade_crop_pct_1980 = "",
    decade_crop_pct_1990 = "",
    decade_crop_pct_2000 = "",
    decade_crop_pct_2010 = "",
    country_1960 = "60s",
    country_1970 = "70s",
    country_1980 = "80s",
    country_1990 = "90s",
    country_2000 = "2000s",
    country_2010 = "2010s"
  ) %>%
  gt_theme_cropc()
```

I like it! There are a few issues, mainly with differing flag sizes because of the way they were uploaded on Wikipedia. We could potentially manually alter these by creating a column in the `flags` dataframe and using the `web_image()` function to transform images based on height / width specified in that but for the sake of this exercise, I'm happy with this. The main thing I'm unhappy with is the column names as I want to do something similar for my second table but unfortunately `cols_label()` does not yet seem to support `match()` regex functions.

Just to demonstrate how I could now make a similar looking table with the function, I thought it'd be interesting to look at Netherlands and the UK in isolation and see where they each rank for every crop. Will remove **bananas**, **rice**, and **soybeans** as neither country export these. This is the data I'll use:

```{r}
uk_nl_ranked <- 
  eu_crop_data %>%
  group_by(decade, crop) %>%
  mutate(decade_crop_pct = ifelse(yield == 0, 0, yield / sum(yield))) %>%
  arrange(desc(decade_crop_pct), .by_group = TRUE) %>% 
  mutate(rank = row_number()) %>% 
  ungroup() %>%
  filter(code %in% c("GBR", "NLD")) %>% 
  mutate(rank = ifelse(yield == 0, 0, rank)) %>%
  left_join(flags, by = "country") %>%
  select(code, decade, rank, flag_url, crop) %>%
  rename("country" = code) %>%
  filter(!crop %in% c("bananas", "rice", "soybeans")) %>%
  pivot_wider(names_from = c(decade, country), values_from = c(rank, flag_url)) %>% 
  select(crop, matches("flag.*60.*NLD"), matches("60_NLD"), matches("flag.*60.*GBR"), matches("60_GBR"),
         matches("flag.*70.*NLD"), matches("70_NLD"), matches("flag.*70.*GBR"), matches("70_GBR"),
         matches("flag.*80.*NLD"), matches("80_NLD"), matches("flag.*80.*GBR"), matches("80_GBR"), 
         matches("flag.*90.*NLD"), matches("90_NLD"), matches("flag.*90.*GBR"), matches("90_GBR"), 
         matches("flag.*2000.*NLD"), matches("2000_NLD"), matches("flag.*2000.*GBR"), matches("2000_GBR"), 
          matches("flag.*2010.*NLD"), matches("2010_NLD"), matches("flag.*2010.*GBR"), matches("2010_GBR"))
```

#### Unaltered table

```{r echo=FALSE}
uk_nl_ranked %>%
  head(3) %>%
  gt()
```

#### Table with function

```{r}
uk_nl_ranked %>%
  gt() %>%
  cols_label(
    flag_url_1960_GBR = "",
    flag_url_1970_GBR = "",
    flag_url_1980_GBR = "",
    flag_url_1990_GBR = "",
    flag_url_2000_GBR = "",
    flag_url_2010_GBR = "",
    flag_url_1960_NLD = "",
    flag_url_1970_NLD = "",
    flag_url_1980_NLD = "",
    flag_url_1990_NLD = "",
    flag_url_2000_NLD = "",
    flag_url_2010_NLD = "",
    rank_1960_GBR = "",
    rank_1970_GBR = "",
    rank_1980_GBR = "",
    rank_1990_GBR = "",
    rank_2000_GBR = "",
    rank_2010_GBR = "",
    rank_1960_NLD = "",
    rank_1970_NLD = "",
    rank_1980_NLD = "",
    rank_1990_NLD = "",
    rank_2000_NLD = "",
    rank_2010_NLD = ""
  ) %>%
  tab_spanner(
    label = "60s",
    columns = 2:5
  ) %>%
  tab_spanner(
    label = "70s",
    columns = 6:9
  ) %>%
  tab_spanner(
    label = "80s",
    columns = 10:13
  ) %>%
  tab_spanner(
    label = "90s",
    columns = 14:17
  ) %>%
  tab_spanner(
    label = "2000s",
    columns = 18:21
  ) %>%
  tab_spanner(
    label = "2010s",
    columns = 22:25
  ) %>%
  tab_header(
    title = md("**The Netherlands and the UK ranked by crop yield ratios**"),
    subtitle = md("0 means *unranked*")
  ) %>%
  gt_theme_cropc()
```


It's definitely not the easiest table to read but can see that it's quick to make tables look similar with a simple function. If it weren't for the drastically different column names, could have included that in the function and reduce the code length drastically. Hopefully in the future `gt` will allow us to use a `match()` regex for column names in which case we could reduce the `cols_label()` parameters to about 3 lines.

Really enjoyed using `gt` and look forward to seeing the tables people have submitted for the RStudio [data contest](https://blog.rstudio.com/2020/09/15/announcing-the-2020-rstudio-table-contest). Definitely going to be implementing some themes at work where we can include things like the brand colours and logo!















